---
title: "Thesis Draft"
output: html_notebook
---

## Introduction
The following R markdown file contains the code to reproduce the cross-sectional
analyses. Data from Baseline and Week 8 are employed separately. The code is
divided in 7 parts:
- Data preparation
- Missing data and multiple imputation
- Exploratory factor analysis (EFA)
- Factor scores exploration
- Optimal number of clusters
- Clustering
- Clusters' interpretation
- 3D Representation

##########################
## Part 1: DATA PREPARATION
##########################

PACKAGE INSTALLATION:

```{r}
#install.packages('readxl') 
#install.packages('mice')
#install.packages('caret')
#install.packages('psych')
#install.packages('cluster')
#install.packages('factoextra')
#install.packages('plotly')
#install.packages('dplyr')
```


NECESSARY LIBRARIES ARE IMPORTED

```{r}
library("readxl") #
library("mice") #
library("caret") #
library("psych") #
library("cluster")    #
library("factoextra") #
library("plotly") #
library("dplyr") #
```


LOADING THE DATASET

The dataset imported contains the demographic information about the patients,
namely:
- Centre: The rheumatology centre that the patient attended
- AgeW104: The age of the patient at the end of the trial (at week 104)
- Gender
- RF: Presence of rheumatoid factor in the serum (yes/no)
- ACPA: Presence of Anti-cyclic citrullinated peptide antibody in the serum (yes/no)
- Eros: Presence of bone erosions (yes/no)

Furthermore, the dataset contains the nine variables employed in the analyses,
assessed at the 10 visits. These are.
- HAQ: Health Assessment Questionnaire
- VAS.Pain: Pain measured on a visual analog scale (VAS)
- TJC28: Tender joint count
- SJC28: Swollen joint count
- CRP: C-reactive protein
- VASASS: Patient global health assessment measured on a VAS
- VASPHYS: Physician global health assessment measured on a VAS
- VASFAT: Fatigue measured on a VAS
- ESR:Erythrocyte sedimentation rate

For further information on the meaning on each variable, please consult
Chapter 2 of the thesis.
```{r}
df = read_excel("CareRAbase.xlsx")
```


DATA EXPLORATION 

```{r}
# Amount of NAs in every columns
colSums(is.na(df[,10:27]))

# Transform the columns containing the variables into numeric 
df[,10:27]=as.data.frame(lapply(df[,10:27],as.numeric))

# Remove the patientID, the weight and height of the patients and any measurements
# after week 8
df = df[,c(2:7,10:27)]
```


################################################
## Part 2: MISSING DATA AND MULTIPLE OUTPUTATION
################################################

MISSINGNESS

Compute the percentage of missingness at the two visits (this will be useful for
the multiple imputation)

```{r}
#Percentage of missing data among the variables
print(paste(round(sum(colSums(is.na(df[,7:24])))/(18*379)*100,1), "% is the 
percentage of missing data among the variables in the cross sectional analyses"))

#Percentage of missing data among the variables
print(paste(round(sum(colSums(is.na(df[,7:15])))/(9*379)*100,1), "% is the 
percentage of missing data among the variables at baseline"))

#Percentage of missing data among the variables
print(paste(round(sum(colSums(is.na(df[,16:24])))/(9*379)*100,1), "% is the 
percentage of missing data among the variables at week 8"))
```


MULTIPLE IMPUTATION 

Multiple imputation is carried out using classification and regression trees as 
conditional model. We impute two dataset given the 1.9% of missing data between
the two analyses. Further information can be found in Section 3.1 of the thesis
```{r}
tempData <- mice(df, m=2, maxit=5, meth='cart', minbucket = 5, seed=500)
```

############################################
## Part 3: EXPLORATORY FACTOR ANALYSIS (EFA)
############################################

COMPUTE THE CORRELATION MATRIX

EFA is carried out on the correlation matrix of the dataset. Since we have two
imputed, complete datasets, we compute the correlation matrix of each and average
them. EFA is then carried out on the average correlation matrix.

```{r}
# Retrieve the complete dataset
p1 = complete(tempData,1)[,(7:24)]
p2 = complete(tempData,2)[,(7:24)]

# Compute the average correlation matrix
correlation = (cor(p1)+cor(p2))/2

# Separate the correlation matrix between the two visits
c_baseline = correlation[(1:9),(1:9)] #correlation of the variables at baseline
c_w8 = correlation[(10:18),(10:18)] #correlation of the variables at week 8
```


EFA WITH PRINCIPAL COMPONENT EXTRACTION

We extract 3 factors and rotate them through promax rotation. Further details 
about the extraction and rotation can be found in Section 3.2 of the thesis

Below there are the factor loadings at baseline. At baseline, the assessment by
the physician has a double loading, on the clinical and the patient-reported 
factors. This is because, at baseline, the assesment of the physician is 
correlated with what the patient thinks and the clinical results.

```{r}
fit_baseline <- principal(c_baseline, nfactors=3, rotate="promax")

print(fit_baseline$loadings, cutoff = 0.3)
```

Below there are the factor loadings at Week 8. No factor loadings are present

```{r}
fit_w8 <- principal(c_w8, nfactors=3, rotate="promax")

print(fit_w8$loadings,cutoff = 0.3)
```


TUCKERS CONGRUENCE COEFFICIENT

Through Tucker's congruence coefficient, we can evaluate the similarity between
multiple factor interpretations, specifically between the factors uncovered at 
the two visits. Given the high coefficients along the diagonal (all 0.99), the
homonymous factors can be considered as equal.

```{r}
factor.congruence(fit_baseline$loadings, fit_w8$loadings)
```

RESCALING OF THE VARIABLES

To obtain the factor scores we need to first average the complete datasets (this
can be done since the differences between the two complete datasets are minimal
and the percentage of missingness was low).
Variables need to be rescaled because they are measured on different ranges.
HAQ for example has a range [0-3], whereas any variable measured on the VAS has
a range [0-100]. All the variables are rescaled to a [0-100] range. Section 3.3
of the thesis contains further information regarding the rescaling of CRP and ESR.

```{r}
#Average the two complete datasets to get the factor values for each observation
complete_df = (p1+p2)/2


#Rescale the dataset so that each variable ranges from 0 to 100
for (col_name in names(complete_df)) {
  if (grepl("HAQ", col_name))
    complete_df[,col_name] = complete_df[,col_name]/3*100
  else if (grepl("JC", col_name))
    complete_df[,col_name] = complete_df[,col_name]/28*100 #Rescaling of both TJC and SJC
  else if (grepl("CRP", col_name)) {
    complete_df[,col_name] = complete_df[,col_name]/20*100 
    complete_df[,col_name] = pmin(complete_df[,col_name], 100)
  }
  else if (grepl("ESR", col_name)) {
    complete_df[,col_name] = complete_df[,col_name]/88*100 
    complete_df[,col_name] = pmin(complete_df[,col_name], 100)
  }
}
```

FACTOR SCORES

We create the factor scores matrix by  multiplied the scaled variables by their
factor loadings and sum them up so that each factor would comprise the variables
that loaded onto it. This is done for both the factor loadings at baseline and 
at week 8. Due to the different number of variables for each factor, each factor 
score was then standardized to a [0–1] scale.


```{r}
# Create the factor matrix
factors = data.frame(matrix(ncol = 0, nrow = 379))

# Patient-reported factor at baseline
factors$PRF1 = (0.885 * complete_df$VASFAT.0) + (0.929 * complete_df$VASASS.0) + (0.964 * complete_df$VAS.pain.0) + (0.594 * complete_df$HAQ.0)
# Clinical factor at baseline
factors$CF1 = (0.962 * complete_df$TJC28.0) + (0.967 * complete_df$SJC28.0) + (0.640 * complete_df$VASPHYS.0)
#Laboratory factor at baseline
factors$LF1 = (0.891 + complete_df$CRP.0) + (0.889 * complete_df$ESR.0)

# Patient-reported factor at week 8
factors$PRF2 = (0.938 * complete_df$VASFAT.8) + (0.902 * complete_df$VASASS.8) + (0.871 * complete_df$VAS.pain.8) + (0.618 * complete_df$HAQ.8)
# Clinical factor at week 8
factors$CF2 = (0.891 * complete_df$TJC28.8) + (0.975 * complete_df$SJC28.8) + (0.677 * complete_df$VASPHYS.8)
# Laboratory factor at week 8
factors$LF2 = (0.838 + complete_df$CRP.8) + (0.853 * complete_df$ESR.8)

#Rescale the factors 
process <- preProcess(as.data.frame(factors), method=c("range"))
factors <- predict(process, as.data.frame(factors))
```

####################################
## Part 4: FACTOR SCORES EXPLORATION
####################################

HISTOGRAMS

Through various exploration techniques we check the distribution of the factors.
First, we compare the distribution at baseline and at week 8 for each of the 
factor scores through histograms.

Patient reported factor:
```{r}
par(mfrow=c(2,1))
par(mar=c(0,5,3,3))

hist(factors$PRF1, breaks = 100, main="" , ylim=c(0,35), ylab="Baseline", xlab="", xaxt="n", las=1 , col="#56B4E9")

par(mar=c(5,5,0,3))
hist(factors$PRF2, breaks = 100, main="" , ylim=c(35,0), ylab="Week 8", xlab="Patient Factor", las=1 , col="#009E73" )
```

Clinical Factor:
```{r}
par(mfrow=c(2,1))
par(mar=c(0.5,5,3,3))

hist(factors$CF1, breaks = 100, main="" , ylim=c(0,80), ylab="Baseline", xlab="", xaxt="n", las=1 , col="#56B4E9")

par(mar=c(5,5,0,3))
hist(factors$CF2, breaks = 100, main="" , ylim=c(80,0), ylab="Week 8" , xlab=" Clinical Factor" , las=1 , col="#009E73" )
```

Laboratory factor:
```{r}
par(mfrow=c(2,1))
par(mar=c(0,5,3,3))

hist(factors$LF1, breaks = 100, main="" , ylim=c(0,35), ylab="Baseline", xlab="", xaxt="n", las=1 , col="#56B4E9")

par(mar=c(5,5,0,3))
hist(factors$LF2, breaks = 100, main="" , ylim=c(35,0), ylab="Week 8", xlab="Laboratory Factor", las=1 , col="#009E73" )
```
COMPARE FACTOR SCORES BETWEEN BASELINE AND WEEK 8

For each factor, we compare the factor scores for each patient between Baseline
and Week 8.

Patient reported factor:
```{r}
ggplot(data = factors, mapping = aes(x = PRF1, y = PRF2)) +
  geom_point() +
  geom_abline(lty = 2) +
  coord_fixed() +
  labs(x = "Baseline", y = "Week 8") +
  theme_minimal() +
  theme(text=element_text(size=20))
```

Clinical factor:
```{r}
ggplot(data = factors, mapping = aes(x = CF1, y = CF2)) +
  geom_point() +
  geom_abline(lty = 2) +
  coord_fixed() +
  labs(x = "Baseline", y = "Week 8") +
  theme_minimal() +
  theme(text=element_text(size=20))
```

Laboratory factor:
```{r}
ggplot(data = factors, mapping = aes(x = LF1, y = LF2)) +
  geom_point() +
  geom_abline(lty = 2) +
  coord_fixed() +
  labs(x = "Baseline", y = "Week 8")+
  theme_minimal() +
  theme(text=element_text(size=20))
```


#####################################
## Part 5: OPTIMAL NUMBER OF CLUSTERS
#####################################

BASELINE

For each visit we employ various techniques to uncover the optimal number of clusters.
For more information about the choice of methods, please refer to Sections 3.4 
of the thesis.

Ultimately, four was chosen as the optimal number of clusters for both analyses.
More information can be found at Section 4.2 of the thesis

```{r}
baseline_factors = factors[,(1:3)]
```


Average Silhouette Method
```{r}
fviz_nbclust(baseline_factors, kmeans, method = "silhouette") +
  ggtitle("") +
  theme(text=element_text(size=20))
```


Gap Statistic Method
```{r}
gap_stat <- clusGap(baseline_factors, FUN = kmeans, nstart = 30, K.max = 10, B = 50)
fviz_gap_stat(gap_stat)+
  ggtitle("") +
 theme(text=element_text(size=20))
```


WEEK 8

```{r}
week8_factors = factors[,(4:6)]
```


Average Silhouette Method
```{r}
fviz_nbclust(week8_factors, kmeans, method = "silhouette") +
  ggtitle("") +
  theme(text=element_text(size=20))
```


Gap Statistic Method
```{r}
set.seed(123)
gap_stat <- clusGap(week8_factors, FUN = kmeans, nstart = 25,
                    K.max = 10, B = 50)

fviz_gap_stat(gap_stat) +
  ggtitle("") +
  theme(text=element_text(size=20))
```

#####################
## Part 6: CLUSTERING
#####################

HKMEANS

Hierarchical K-means clustering is applied to both visits, uncovering four 
clusters at each time point.

```{r}
baseline = hkmeans(baseline_factors, 4)

week8 = hkmeans(week8_factors, 4)
```

CLUSTER ALIGNMENT

By analyzing the centroids of the four clusters from the two visits we discovered 
which clusters could be aligned. Since the clustering did not output the clusters
in the same order across the two visits, we need to align these.
At baseline, the clusters found are labelled: "mildly symptomatic", "unmet needs", 
"non-inflammatory burden", and "vulnerable".
At week 8, the clusters found are labelled: "mildly symptomatic", "unmet needs", 
"non-inflammatory burden", and "inflammatory burden".
We align the clusters so that they are in the same order as reported here above.

We create a function named swap_rows. This function swap the rows of the centroids,
so that they are shown in the order displayed above.

```{r}
swap_rows <- function(df, index1, index2) {
  temp <- df[index1, ]
  df[index1, ] <- df[index2, ]
  df[index2, ] <- temp
  return(df)
}

# These are the centroids as output from the HKmeans algorithm
centroids_bl = baseline$centers
centroids_w8 = week8$centers

# We swap the rows so that they have the same order as displayed above
centroids_bl = swap_rows(centroids_bl,1,3)
centroids_bl = swap_rows(centroids_bl,4,3)
centroids_w8 = swap_rows(centroids_w8,1,2)
centroids_w8 = swap_rows(centroids_w8,3,4)
```

PATIENT ALIGNMENT

We have aligned the clusters' centroids, but the patients are still assigned to 
the cluster number from before the alignment. The membership of the patients is
stored in the "baseline\$cluster" and "week8\$cluster". These vectors contain
a series of 379 values between 1 and 4 indicating the patient's cluster membership. 
We need to align these based on the previous clusters alignment.

```{r}
transform_vector <- function(input_vector, mapping) {
  output_vector <- input_vector
  for (from_value in names(mapping)) {
    to_value <- mapping[from_value]
    output_vector[input_vector == as.numeric(from_value)] <- as.numeric(to_value)
  }
  return(output_vector)
}

# Mapping for baseline
custom_mapping_bl <- c("1" = "4", "3" = "1", "4" = "3") 
# Mapping for Week 8
custom_mapping_w8 <- c("2" = "1", "1" = "2", "3" = "4", "4" = "3") 

# Updated vectors containing the patients cluster memberships
cluster_bl <- transform_vector(baseline$cluster, custom_mapping_bl)
cluster_w8 <- transform_vector(week8$cluster, custom_mapping_w8)
```


PATIENT-CLUSTER MEMBERSHIP

Through the table function we can check how the membership of the patients to the
clusters changes at the two time points. The leftmost column contains the four
different clusters from Baseline, while the top row specifies the four clusters
from Week 8.

```{r}
table(cluster_bl, cluster_w8)
```

###################################
## Part 7: CLUSTERS' INTERPRETATION
###################################

To determine the clinical relevance of the cluster solutions, we examined 
socio-demographic and health variables, as well as treatment strategies.

For each cluster we checked:
- Mean age
- Percentage of women
- Percentage of patients that have been classified as low-risk
- Percentage of patients that displayed comorbidities at baseline
- Distribution of patients across the clusters for each treatment strategy

We load another file containing extra information on the patients

```{r}
group = read_excel("Interpretation.xlsx")
```


We create a dataframe that contains for each patients:
- Patient ID
- The factor scores at baseline (1) and week8 (2)
- The clusters they belonged to at baseline and at week 8
- The demographics employed at the beginning of the analysis (gender, age, etc.)
- The treatment group
- The risk group
- Presence of comorbidities

```{r}
clusters = cbind(group[,1], factors, cluster_bl, cluster_w8, df[,1:6], group[,2:5])
clusters$`risk group` = as.factor(clusters$`risk group`)
clusters$Comorbidities = as.factor(clusters$Comorbidities)
clusters$treatmentgroup = as.factor(clusters$treatmentgroup)
```


The following functions give a summary of a specific variable for each for the
clusters. The function "calculate_stats_by_group" gives the mean and standard
deviation of a certain variable for each of the clusters. It can be employed for
numerical variables (such as age, but also the factor scores at the two visits)

```{r}
calculate_stats_by_group <- function(dataframe, value_col, group_col) {
  result <- tapply(dataframe[[value_col]], dataframe[[group_col]], function(x) c(mean = mean(x), sd = sd(x)))
  print(result)
}

# Examples
calculate_stats_by_group(clusters, "AgeW104", "cluster_bl")
calculate_stats_by_group(clusters, "PRF1", "cluster_bl")
calculate_stats_by_group(clusters, "CF2", "cluster_w8")
```


The function "generate_summary" provides a distribution of patients across the 
clusters for a certain categorical variable. It can be employed for factors such
as the gender, risk group or comorbidities. 

```{r}
generate_summary <- function(dataframe, categorical_col, based_on_col) {
  summary_table <- table(dataframe[[based_on_col]], dataframe[[categorical_col]])
  print(summary_table)
}

# Examples
generate_summary(clusters, "risk group", "cluster_bl")
generate_summary(clusters, "Gender", "cluster_bl")
generate_summary(clusters, "Comorbidities", "cluster_w8")
```
############################
## Part 8: 3D REPRESENTATION
############################

We provide a 3D representation of the patients scores at each time visits, colored
by their cluster memberships

Baseline
```{r}
p <- plot_ly(clusters, x=~PRF1, y=~CF1, z=~LF1, color=as.factor(cluster_bl), colors = c("#02e066", "#E7B800", "#FC4E07", "#030103")) %>%
     add_markers(size=1.5)
print(p)
```


Week 8
```{r}
p <- plot_ly(clusters, x=~PRF2, y=~CF2, z=~LF2, color=as.factor(clusters$cluster_w8), colors = c("#02e066", "#E7B800", "#FC4E07", "#af2bd4")) %>%
     add_markers(size=1.5)
print(p)
```


```{r}
cl_2 = clusters
cl_2$cluster_w8[cl_2$cluster_bl == 1 & cl_2$cluster_w8 == 2] = 5
cl_2$cluster_w8[cl_2$cluster_bl == 2 & cl_2$cluster_w8 == 3] = 6


p <- plot_ly(cl_2, x=~PRF2, y=~CF2, z=~LF2, color=as.factor(cl_2$cluster_w8), colors = c("#02e066", "#E7B800", "#FC4E07", "black", "purple", "royal blue")) %>%
     add_markers(size=1.5)
print(p)
```


Variable exploration


```{r}
hist(complete_df$CRP.0, breaks = 30)
boxplot(complete_df$ESR.0, xlab="ESR")

ggplot(complete_df, aes(x = "", y = CRP.0)) + 
  geom_jitter(width = 0.2) +
  geom_boxplot(alpha = 1, width = 0.4,  outlier.color = "transparent")
```

